Technique for \earlywarning\ filtering for compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-29 introduction.tex
Worked around bug in hyperref package
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1195 ccdd05d5-56c1-461a-bf98-85f6876523b4
added references and blurb on IIR
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1196 ccdd05d5-56c1-461a-bf98-85f6876523b4

\subsection{\EM{} counterparts}

Compact binary coalescence (\CBC) is thought to be a
mechanism for short gamma-ray bursts (short \GRB{}s) \citep{Lee:2005, nakar07}.
In this scenario, prompt gamma-ray emission arises on the accretion timescale
of a central compact object formed after the merger, $0.1$--$1.0$\,s
\citep{Janka1999}.  Interestingly, roughly a quarter to half of observed short
\GRB{}s also exhibit extended emission of $30$--$100$\,s in duration beginning
$\sim$$10$\,s after the \GRB{} and carrying comparable fluence to the initial
outburst.  Potential explanations for the emission are delayed fall-back
accretion caused by $r$-process heating \citep{Metzger2010} or the formation of
a proto-magnetar that interacts with
ejecta \citep{Bucciantini2011}.  A proto-magnetar may
form from accretion-induced collapse or the merger of two \textsc{ns}s.  An
accompanying \GW{} observation would confirm the latter.

@@

In October 2010, \LIGO{}\footnote{\url{http://www.ligo.org/}} completed its
sixth science run (S6) and Virgo\footnote{\url{http://www.ego-gw.it/}}
completed its third science run (VSR3).  While both \LIGO{} detectors and Virgo
were operating, several all-sky detection pipelines operated in a low-latency
configuration to send astronomical alerts, namely \textsc{mbta}, Coherent
WaveBurst, and Omega \citep{HugheyGWPAW2011, S6lowlatency}. The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for \EM{} followup to several telescopes; Swift,
\textsc{rotse}, \textsc{tarot}, and Zadko \citep{kanner2008, HugheyGWPAW2011}
imaged some of the most likely sky locations.  \textsc{mbta} achieved the best
\GW{} trigger-generation latencies of 2--5 minutes.

@@

This work will focus on reducing the latency of trigger production.  Realizing
advance detection of \CBC{}s will require striking a balance between latency
and throughput of filtering \GW\ data. \CBC{} searches consist of banks of
matched filters, or cross-correlations between the data stream and a bank of
nominal ``template'' signals.  There are many different implementations of
matched filters, but most have high throughput at the cost of high latency, or
low latency at the cost of low throughput.  The former are epitomized by the
overlap-save algorithm
for frequency domain (\FD) convolution, currently the preferred method in \GW{}
searches.  The most obvious example of the latter is the time domain (\TD)
convolution, which is latency-free.  However, its computational complexity is
quadratic in the length of the templates, so it is prohibitively expensive for
long templates.

>>>>

\subsection{\EM\ counterparts}

Compact binary coalescence (\CBC) events are thought to be a
mechanism for short gamma-ray bursts (short \GRB{}s) \citep{Lee:2005, nakar07}.
In this scenario, prompt gamma-ray emission arises on the accretion timescale
of a central compact object formed after the merger, $0.1$--$1.0$\,s
\citep{Janka1999}.  Interestingly, roughly a quarter to half of observed short
\GRB{}s also exhibit extended emission of $30$--$100$\,s in duration beginning
$\sim$$10$\,s after the \GRB{} and carrying comparable fluence to the initial
outburst.  Potential explanations for the emission are delayed fall-back
accretion caused by $r$-process heating \citep{Metzger2010} or the formation of
a proto-magnetar that interacts with
ejecta \citep{Bucciantini2011}.  A proto-magnetar may
form from accretion-induced collapse or the merger of two \textsc{ns}s.  An
accompanying \GW{} observation would confirm the latter.

@@

In October 2010, \LIGO{}\footnote{\url{http://www.ligo.org/}} completed its
sixth science run (S6) and Virgo\footnote{\url{http://www.ego-gw.it/}}
completed its third science run (VSR3).  While both \LIGO{} detectors and Virgo
were operating, several all-sky detection pipelines operated in a low-latency
configuration to send astronomical alerts, namely \textsc{mbta}, Coherent
WaveBurst, and Omega \citep{HugheyGWPAW2011, S6lowlatency, S6lowlatency2, S6lowlatency3, S6lowlatency4}. The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for \EM{} followup to several telescopes; Swift,
\textsc{rotse}, \textsc{tarot}, and Zadko \citep{kanner2008, HugheyGWPAW2011}
imaged some of the most likely sky locations.  \textsc{mbta} achieved the best
\GW{} trigger-generation latencies of 2--5 minutes.

@@

This work will focus on reducing the latency of trigger production.  Realizing
advance detection of \CBC{}s will require striking a balance between latency
and throughput of filtering \GW\ data. \CBC{} searches consist of banks of
matched filters, or cross-correlations between the data stream and a bank of
nominal ``template'' signals.  There are many different implementations of
matched filters, but most have high throughput at the cost of high latency, or
low latency at the cost of low throughput.  The former are epitomized by the
overlap-save algorithm for frequency domain (\FD) convolution, currently the
preferred method in \GW{} searches.  The most obvious example of the latter is
the time domain (\TD) convolution, which is latency-free.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.  We note that \citep{shaunIIR} have
explored an infinite-impulse-response (IIR) decomposition of \CBC\ waveforms as
a means to reduce the computational complexity of \TD\ filtering.  They
demonstrate a factor of 50 improvement in floating point operations per second
over the standard \TD\ method.  We will significantly improve upon that number.
